First steps in synthetic algebraic geometry

### 2017-02-07 slides-warwick2017.tex
Final todo items for tomorrow

%\usepackage{mathtools}
\newcommand{\vcentcolon}{:}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows}
\hypersetup{colorlinks=true}

@@

\usefonttheme{serif}
%\usepackage[T1]{fontenc}
%\usepackage{libertine}

@@

\begin{frame}\frametitle{Generic freeness}
  % XXX: enlarge diagram
  \vspace*{-1em}
  \[ \xymatrix{
    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
    R \ar[r]_{\text{of finite type}} & S
  } \]
  If $R$ is reduced ($x^n = 0 \Rightarrow x = 0$), there is a \hil{dense open}~$U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
  \subseteq U$,
  \begin{itemize}
    \item $S[f^{-1}]$ and $M[f^{-1}]$ are free modules over $R[f^{-1}]$,
    \item $R[f^{-1}] \to S[f^{-1}]$ is of finite presentation, and
    \item $M[f^{-1}]$ is finitely presented as a module over~$S[f^{-1}]$.
  \end{itemize}

@@

  \begin{block}{Basic look and feel}
  \begin{itemize}
    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
    points} of~$X$.
    \item Internally, $\affl$ (given by $\affl(X) = \O_X(X)$)
    looks like a field:
    \[ \Zar(S) \models \forall x\?\affl\_ x \neq 0 \Longrightarrow \speak{$x$ invertible} \]
  \end{itemize}
  \end{block}

@@

  \item $\RelSpec_S A = \Hom_{\mathrm{Alg}(\affl)}(A, \affl) =
  \text{set of $\affl$-rational points of $A$}.$

  \item $TX = \Hom(\Delta, X)$,
  where~$\Delta = \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$.

  \item An $\affl$-module~$E$ is quasicoherent if and only if
  \[ E \otimes_\affl A \longrightarrow \Hom(\Spec A, E) \]
  is an isomorphism for all finitely presented $\affl$-algebras $A$.

  \item In particular, any map $\affl \to \affl$ is given by a polynomial.

@@

  \item A subset $U \subseteq X$ is qc-open if and only if for any $x : X$
  there exist $f_1,\ldots,f_n \subseteq \affl$ such that
  \[ x \in U \quad\longleftrightarrow\quad \exists i\_ \text{$f_i$ is invertible}. \]
  \end{itemize}
\end{frame}

@@

\subsection{A strong Kock--Lawvere axiom}

\begin{frame}\frametitle{A strong Kock--Lawvere axiom}
  \begin{itemize}
    \item The affine line fulfils the axiom
      \[ \Zar(S) \models \speak{every function~$\affl \to \affl$ is a
      polynomial}. \]
    More precisely, the canonical morphism
    \[ \affl[T] \longrightarrow \Hom(\Hom_{\mathrm{Alg}(\affl)}(\affl[T],\affl), \affl) \]
    is an isomorphism.
    \item More generally, for any~$\affl$-algebra~$\A$ induced by a
    quasicoherent~$\O_S$-algebra, the canonical morphism
    \[ \A \longrightarrow \Hom(\Hom_{\mathrm{Alg}(\affl)}(\A,\affl), \affl) \]
    is an isomorphism.
  \end{itemize}
\end{frame}

@@

\subsection{Comparing the little and the big toposes}

\begin{frame}\frametitle{Comparing the little and the big toposes}
  \begin{itemize}
  \item There is a local geometric morphism~$\Zar(S) \to \Sh(S)$.

  \item From the point of view of~$\Sh(S)$, the big Zariski topos is $\Zar(\O_S | \O_S)$,
  the classifying topos of local~$\O_S$-algebras which are local
  over~$\O_S$.

  \item From the point of view of~$\Zar(S)$, the little Zariski topos is the
  largest subtopos where $\flat \affl \to \affl$ is bijective.
  \begin{align*}
    (\flat \affl)(X \xrightarrow{\mu} S) &= (\mu^{-1} \O_S)(X) \\
    \affl(X \xrightarrow{\mu} S) &= \O_X(X)
  \end{align*}
  \end{itemize}
\end{frame}

@@

\begin{frame}\frametitle{Semi-open and open tasks}
  \begin{itemize}
  \item Characterise quasicoherence in the big Zariski topos.
  \item Understand how to work with~${\flat} \dashv \sharp$.
  \item Do cohomology in the little Zariski topos; exploit that higher direct
  images look like ordinary sheaf cohomology from the internal point of view.
  \item Do cohomology in the big Zariski topos.
  \item Understand more subtoposes of the big Zariski topos.
  \item Derive suitable axioms for synthetic algebraic geometry.
  \end{itemize}

>>>>

\usepackage{mathtools}
%\newcommand{\vcentcolon}{:}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows}
\hypersetup{colorlinks=true}

@@

\usefonttheme{serif}
\usepackage[T1]{fontenc}
\usepackage{libertine}

@@

\begin{frame}\frametitle{Generic freeness}
  \vspace*{-1.6em}
  \[ \xymatrixcolsep{3pc}\xymatrixrowsep{3pc}\xymatrix{
    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
    R \ar[r]_{\text{of finite type}} & S
  } \]
  If $R$ is reduced ($x^n = 0 \Rightarrow x = 0$), there is a \hil{dense open}~$U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
  \subseteq U$,
  \begin{itemize}
    \item $S[f^{-1}]$ and $M[f^{-1}]$ are free modules over $R[f^{-1}]$,
    \item $R[f^{-1}] \to S[f^{-1}]$ is of finite presentation, and
    \item $M[f^{-1}]$ is finitely presented as a module over~$S[f^{-1}]$.
  \end{itemize}
  \pause

@@

  \begin{itemize}
    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
    points} of~$X$.
    \item Internally, $\affl$ (given by $\affl(X) = \O_X(X)$)
    looks like a field:
    \[ \Zar(S) \models \forall x\?\affl\_ x \neq 0 \Longrightarrow \speak{$x$ invertible} \]
  \end{itemize}

@@

  \item \ \\[-1.2em]\mbox{$\RelSpec_S R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
  \text{set of $\affl$-rational points of $R$}.$}

  \item $TX = \Hom(\Delta, X)$,
  where~$\Delta = \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$.
  \pause

  \item An $\affl$-module~$E$ is quasicoherent if and only if
  \[ E \otimes_\affl R \longrightarrow \Hom(\Spec R, E) \]
  is an isomorphism for all finitely presented $\affl$-algebras $R$.

  \item In particular, any map $\affl \to \affl$ is given by a polynomial.
  \pause

@@

  \item A subset $U \subseteq X$ is qc-open if and only if for any $x : X$
  there exist $f_1,\ldots,f_n \subseteq \affl$ such that
  \[ x \in U \quad\Longleftrightarrow\quad \exists i\_ \text{$f_i$ is invertible}. \]
  \end{itemize}
\end{frame}

% Add two bigger case studies:
% * Construction of the Grassmannian
% * f : Delta --> X such that f(0) in U implies that f factors over U
%
% And add the following smaller things:
% * qc, sep, qs
% * universally closed
% * surjective

@@

@@

@@

\begin{frame}\frametitle{Semi-open and open tasks}
  \begin{itemize}
  \item Do cohomology in the little Zariski topos; exploit that higher direct
  images look like ordinary sheaf cohomology from the internal point of view.
  \item Do cohomology in the big Zariski topos.
  \item Understand more subtoposes of the big Zariski topos.
  \item Derive suitable axioms for synthetic algebraic geometry.
  \end{itemize}
